Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-11 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
Edits to sec 5.3

Roughly similar or even worse behavior was found for thin plate regression splines used in the experiments, with serious challenges with computation time when doing inference for spline models for $D=3$ and with more than 10 basis functions, or even for $D=2$ and more than 40 basis functions (see Figures~B.3-right and C.3 in the online material).

@@

\subsection{Case study of simulated data for $2$D and $3$D functions}\label{sec_caseIII}

In this example we apply the diagnostic tool to fit and diagnose two different data sets, one data set simulated from a bivariate (D=$2$) function and another data set simulated from a $3$-dimensinal (D=$3$) function. Furthermore, we present results about the computational time required to fit the HSGP model in $2$D, $3$D and $4$D input spaces, with different sized data sets and as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation.

The $2$D and $3$D synthetic datasets consist of $n=200$ and $n=1000$ ($i=1,\dots,n$) single draws from GP priors with input values $\bm{x}_i \in \{[-1,1],\dots,[-1,1]\} \subset {\rm I\!R}^D$, in a D-dimensional input space. A squared exponential covariance function, with hyperparameters marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, for the first input dimension, $\ell_2=0.3$, for the second input dimension, and $\ell_3=0.4$, for the third input dimension, is used for the GP priors. Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws to form the final noisy sets of observations.

In the HSGP models with $2$ and $3$ input dimensions, the underlying functions $f(\bm{x})$ to the data are approximated as in eq. (13)% (\ref{eq_approxf_multi})
, with the $D$-dimensional squared exponential spectral densities $S$ as in eq. \eqref{eq_specdens_inf}, and the $D$-vectors of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations \eqref{eq_eigenfunction_multi} and \eqref{eq_eigenvalue_multi}, respectively.

@@

The HSGP models are fitted following the recommended iterative steps as explained in Section~\ref{sec_user_guide}, where in each iteration the diagnosis is applied on every dimensions and, for each one of the dimensions, the parameters $c$, $m$, minimum $\ell$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ are updated by equations in Section~\ref{sec_num_equations}. The values $0.5$, $1$, and $1$ are set as initial values for the minimum lengthscales $\ell_1$, $\ell_2$ and $\ell_3$, for the first, second and third dimensions, respectively. In order to be as efficient as possible in these multivariate cases, after the diagnostic generated the first \textit{true} for a certain dimension, its corresponding $m$ is updated by increasing the $m$ of the previous iteration by only 2 additional basis functions and, after the second \textit{true}, $m$ is no longer increased. The full diagnosis process is applied until two \textit{trues} are achieved for each dimension.

Table~\ref{tab_caseIII_2D} and \ref{tab_caseIII_3D} contain the iterative steps to fit and diagnose the $2$D and $3$D data sets, respectively. 
%These tables contain the values for the parameters $\ell$, $c$, $m$, the estimated $\hat{\ell}$, the diagnostic $\hat{\ell} + 0.01 \geq \ell$ for each dimension, and the RMSE compute with both the data and the GP as the reference for every iterative steps of the fitting process. 
The minimum requirements to fit the models were easily achieved by performing 4 iterations: from a $1^{st}$ iteration that uses few multivariate basis functions $m^*$ ($m^*=m_1(=6) \times m_2(=6)=36$ and $m^*=m_1(=6) \times m_2(=6) \times m_3(=6)=216$ for the $2$D and $3$D data sets, respectively) to the $4^{th}$ iteration that uses the minimum multivariate basis functions ($m^*=m_1(=22) \times m_2(=11)=242$ and $m^*=m_1(=21) \times m_2(=10) \times m_3(=12)=2520$ for the $2$D and $3$D data sets, respectively) required to accurately fit these notably wiggly data sets. Without using our recommended guidelines and diagnostics, it could have taken user several trials to get the optimal solution and to spend a significant amount of time, as using a larger number of basis functions in multivariate cases increases time of computation drastically.

@@

Figure \ref{ch5_fig16_errors_exII} shows the difference functions between the $2$D data-generating function, from where the dataset was drawn, and the GP, HSGP and spline-based model solutions, respectively. The $2$D splines-based model was fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}. For the spline-based model, $40$ knots in each dimension were used. Figure \ref{ch5_fig16_errors_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$ and number of univariate basis functions $m$, for the HSGP, and knots, for the $2$D spline-based model. It can be seen an accurate approximation of the HSGP and exact GP to the $2$D data-generative function. However, the performance of the spline-based model is significantly worse.

>>>>

We observed roughly similar or even worse behavior for the thin plate regression splines used in the experiments, with serious challenges with computation time when doing inference for spline models for $D=3$ and with more than 10 basis functions, or even for $D=2$ and more than 40 basis functions (see Figures~B.3-right and C.3 in the online material).

@@

\subsection{Case study: Simulated data for $2$D and $3$D functions}\label{sec_caseIII}

In this case study, we apply the diagnostic tool to fit and diagnose two different data sets, one data set simulated from a bivariate (D=$2$) function and another data set simulated from a $3$-dimensinal (D=$3$) function. Furthermore, we present results or the computational time required to fit the HSGP model in $2$D, $3$D and $4$D input spaces, with different sized data sets and as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation.

The $2$D and $3$D synthetic datasets consist of $n=200$ and $n=1000$ ($i=1,\dots,n$) single draws from GP priors with input values $\bm{x}_i \in \{[-1,1],\dots,[-1,1]\} \subset {\rm I\!R}^D$, in a D-dimensional input space. We use a GP prior with a squared exponential covariance function with hyperparameters marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, $\ell_2=0.3$, and $\ell_3=0.4$, where $\ell_i$ is the length scale for the $i$'th dimension. I.I.D. Gaussian noise with standard deviation $\sigma=0.2$ was added to the GP draws to form the final noisy sets of observations.

In the HSGP models with $2$ and $3$ input dimensions, the underlying functions $f(\bm{x})$ are approximated as in eq. (13)% (\ref{eq_approxf_multi})
, with the $D$-dimensional squared exponential spectral densities $S$ as in eq. \eqref{eq_specdens_inf}, and the $D$-vectors of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations \eqref{eq_eigenfunction_multi} and \eqref{eq_eigenvalue_multi}, respectively.

@@

The HSGP models are fitted following the recommended iterative steps as explained in Section~\ref{sec_user_guide}, where the diagnosis is applied on every dimensions. For each dimension, the parameters $c$, $m$, minimum $\ell$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ are updated using the equations in Section~\ref{sec_num_equations}. The values $0.5$, $1$, and $1$ are set as initial values for the minimum lengthscales $\ell_1$, $\ell_2$ and $\ell_3$, respectively. In order to be as efficient as possible, after the diagnostic generated the first \textit{true} for a certain dimension, its corresponding $m$ is updated by increasing the $m$ of the previous iteration by only 2 additional basis functions and, after the second \textit{true}, $m$ is no longer increased. The full diagnosis process is applied until two \textit{trues} are achieved for each dimension.

Table~\ref{tab_caseIII_2D} and \ref{tab_caseIII_3D} contain the iterative steps to fit and diagnose the $2$D and $3$D data sets, respectively. 
%These tables contain the values for the parameters $\ell$, $c$, $m$, the estimated $\hat{\ell}$, the diagnostic $\hat{\ell} + 0.01 \geq \ell$ for each dimension, and the RMSE compute with both the data and the GP as the reference for every iterative steps of the fitting process. 
The minimum requirements to fit the models were easily achieved by performing 4 iterations: from a $1^{\text{st}}$ iteration that uses few multivariate basis functions $m^*$ ($m^*=m_1(=6) \times m_2(=6)=36$ and $m^*=m_1(=6) \times m_2(=6) \times m_3(=6)=216$ for the $2$D and $3$D data sets, respectively) to the $4^{\text{th}}$ iteration that uses the minimum multivariate basis functions ($m^*=m_1(=22) \times m_2(=11)=242$ and $m^*=m_1(=21) \times m_2(=10) \times m_3(=12)=2520$ for the $2$D and $3$D data sets, respectively) required to accurately fit these notably wiggly data sets. Without using our recommended guidelines and diagnostics, it could have taken user several trials to get the optimal solution and to spend a significant amount of time, as using a larger number of basis functions in multivariate cases increases time of computation drastically.

@@

Figure \ref{ch5_fig16_errors_exII} shows the difference  between the true underlying data-generating function and the posterior mean for GP, HSGP and spline-based model solutions, respectively. The splines-based model was fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}. For the spline-based model, $40$ knots in each dimension were used. Figure \ref{ch5_fig16_errors_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$ and number of univariate basis functions $m$, for the HSGP, and knots, for the $2$D spline-based model. It can be seen that the posterior mean for HSGP and exact GP accurately represents the true data-generating function. In contrast, the accuracy of the spline-based model is significantly worse.
